---
title: "Processing experts shortlists"
author: 
  - Damiano Oldoni
  - Tim Adriaens
date: "`r Sys.Date()`"
output:
  html_document:
  toc: true
toc_depth: 3
toc_float: true
number_sections: true
theme: yeti
df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---
  
```{r global_options, include = FALSE}
knitr::opts_chunk$set(root.dir = "../",
                      warning=FALSE, 
                      message=FALSE)
```

# Load libraries

```{r load_libraries}
# Work with googlesheets
library(googlesheets)

# Tidyverse packages
library(dplyr)
library(stringr)
library(readr)
library(tidyr)
library(purrr)
library(magrittr)
```

# Import spreadsheets

The experts receive a spreadsheet with a *longlist* of alien or possible future aliens. The experts are invited to review it by adding additional species on it. In the column `checklist` they should write `expert check`. Each of them is  invited to select around 10-20 species. Individual spreadsheets are made which are not globally shared with all experts community. At the moment this strategy is being tested within the vertebrates group.

The Google spreadsheets are called  in the following way: `ias_UKOT_vertebrates_firstname_surname.tsv`. It contains two sheets, `LONGLIST` and `SHORLIST`. The shortlist is structured as follows:

|species|country|remarks (optional)|
| --- | --- | --- |
|   example1   |   KY, AI, BM   |   remark1   |
|   example2   |   AI, BM, MS, TC, VG, KY|   remark2   |
|   example3   |   TC   |   ...   |
|    |    |    |

The columns `species` and  `country` are mandatory while `remarks` is optional.
Countries should be comma separated values as shown in the table above and should contain one or more of the 6 Caribbean UKOTs.

We select the spreadsheets to import:

```{r select_spreadsheets}
experts <- c("Elena_Tricarico",
             "Fred_Burton",
             "Ian_Winfield",
             "Jane_Haakonsson_Sophie_O’Hehir",
             "Latisha_Martin",
             "Mark_Outerbridge",
             "Nancy_Pascoe",
             "Rose-Smyth_Christine",
             "Jason_Berry",
             "Bryan_Manco",
             "Tim_Adriaens")
gss_experts <- str_c("ias_UKOT_vertebrates_", experts, ".tsv")
my_sheets <- gs_ls()
experts_ss <- my_sheets %>% 
  filter(sheet_title %in% gss_experts)
experts_ss <- map_dfr(unique(experts_ss$sheet_title), 
                      function(x) experts_ss %>% 
                        filter(sheet_title == x) %>%
                        filter(updated == max(updated)))
# vector experts present in Google Drive and ordered as such
experts_from_gssheets <- unlist(
  map(
    experts_ss$sheet_title, ~ str_split(
      ., pattern = "ias_UKOT_vertebrates_")[[1]][2]
))
experts_from_gssheets <- str_sub(experts_from_gssheets, 
                                 end = -(str_length(".tsv")+1))
experts_ss$sheet_title
```

We import the experts' shortlists by these spreadsheets:

```{r import_spreadsheets, results= "hide"}
title_ssheets <- map(experts_ss$sheet_title, ~ gs_title(.))
dfs_experts <- map(
  title_ssheets, ~ gs_read(ss = ., ws = "SHORTLIST")
)
```

We provide names to this list of dataframes. We name them by expert name:

```{r name_by_expert}
names(dfs_experts) <- experts_from_gssheets
```


We save the shortlists in the directory `./data/raw/`:

```{r save_raw_shortlists, results = "hide"}
map(names(dfs_experts), ~ write_tsv(dfs_experts[[.]], 
                                    path = paste0("../data/raw/ias_UKOT_vertebrates_", ., ".tsv")))
```

An example from the shortlist of one expert:

```{r explore}
dfs_experts$Tim_Adriaens
```


Tidy the dataframes (one row per species per country):

```{r tidy_dfs}
# remove any whitespace experts could ever have introduced
dfs_experts <- map(dfs_experts, function(x) x %>%
                     mutate(UKOTs = str_replace_all(UKOTs, 
                                                pattern = " ",
                                                replace = "")))
dfs_experts <- map(dfs_experts, 
                   function(x) x %<>% 
                     separate_rows(UKOTs, sep = ","))
# add name experts
dfs_experts <- map(names(dfs_experts),
                   function(x) dfs_experts[[x]] %<>%
                     mutate(expert = x))
```

We can now combine the information provided by all experts:

```{r combine_dfs_experts}
combined_dfs <- bind_rows(dfs_experts)
combined_counts_species <- combined_dfs %>% group_by(species, 
                                                     UKOTs) %>%
  summarize(n = n(), 
            experts = toString(na.omit(expert)),
            remarks = toString(na.omit(`remarks (optional)`))) %>%
  ungroup() %>%
  filter(!species == "Dummy Dummy") %>%
  distinct() %>%
  mutate(experts = str_replace_all(experts, 
                                   pattern = " ",
                                   replace = ""))
```

Notice that we *summarized* preserving the names of the experts. It will be a very useful information in order to discuss scoring among experts. First, we have to check the consistency between the merged shorlist and longlist. It can happen that an expert mistypes a species name or adds a species in his shortlist without first adding it in the longlist.

```{r check_consistency_merged_shortlist_longlist}
source('./check_completeness_longlist.R')
check_completeness_longlist(
  longlist_filename = "ias_UKOT_vertebrates_invertebrates.tsv", 
  merged_species_df = combined_counts_species, 
  ws_longlist = "longlist", 
  row_number_longlist = 1,
  col_name_species_longlist = "species")
```

Species in column `species` have to be checked and in case added to longlist. Again, the column `experts` can help in case some doubts arise.

Which species are present in the combined shorlist?

```{r general_species_list}
species_list <- combined_counts_species %>%
  arrange(desc(n)) %>%
  distinct(species)
species_list
```

To a better overview, let's spread such list by UKOTs:

```{r spread_table}
merged_species_list_spread <- spread(combined_counts_species, 
                                     key = UKOTs,
                                     value = n)
merged_species_list_spread
```

And save it:

```{r save_spread_table}
write_tsv(merged_species_list_spread, path = "../data/output/merged_species_list_spread.tsv")
```

It could be also important to have an overview of species combined with the number of experts who judge them present in the horizon. In case the same species has different number of assessors for different UKOTs, take the maximum.

```{r species_number_experts}
overview_species <- combined_counts_species %>% 
  group_by(species) %>% 
  summarize(n_experts = max(n)) %>%
  arrange(desc(n_experts))
overview_species
```

Let's save it in `../data/output/first_round_as_UKOTs.tsv`:

```{r write_general_species_list}
write_tsv(overview_species,
          path = "../data/output/overview_species_number_experts.tsv")
```

# Focus on shortlists

Experts should now start to set scores for species in the combined shortlist. In order to help them in doing it, we provide them a selection of the unified checklist with GBIF number of occurrences based on the species in the combined shorlist. 

First we have to get the most updated version of the longlist which has been saved as a googlesheet `workflow_retrieve_GBIF_checklist_dplyr.Rmd`:

```{r get_most_udpated_longlist, results = "hide"}
longlist_ssheet <- "ias_UKOT_vertebrates_invertebrates"
my_sheets <- gs_ls()
google_ssheet <- my_sheets %>% 
  filter(str_detect(my_sheets$sheet_title, 
                    pattern = longlist_ssheet))
google_ssheet %<>%
  filter(updated == max(google_ssheet$updated)) # take the most recent version
longlist_vert <- gs_key(google_ssheet$sheet_key) %>% 
  gs_read(ws = "longlist")
```

We select the species which are in the shortlists:

```{r}
combined_shortlist_occ <- longlist_vert %>%
  filter(species %in% overview_species$species)
combined_shortlist_occ
```

We save it as an output in order to share it with experts for starting scoring:

```{r save_checklist_occ_filtered_by_shortlist}
write_tsv(
  combined_shortlist_occ, 
  path = "../data/output/filter_combined_shortlist_from_unified_checklist.tsv")
```
