---
title: "Building a Caribbean IAS checklist of chordata"
author: 
- Damiano Oldoni
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
    theme: yeti
    df_print: paged
knit: (function(input_file, encoding) { rmarkdown::render(input_file, encoding = encoding, output_file = paste0("../docs/",sub(".Rmd", ".html", basename(input_file))))})
---

```{r global_options, include = FALSE}
knitr::opts_chunk$set(root.dir = "../",
                      warning=FALSE, 
                      message=FALSE)
```

Install packages:

```{r install_packages}
devtools::install_github("trias-project/trias")
devtools::install_github("ropensci/rgbif")
```

Load libraries:

```{r load_libraries}
# Tidyverse packages
library(dplyr)
library(tidyr)
library(magrittr)
library(readr)
library(stringr)
library(purrr)
# Other packages
library(janitor)
library(lazyeval)
library(rgbif)
library(trias)
library(googlesheets)
```

Set file paths:

```{r set_file_paths}
# All downloads we've triggered are added to this file:
downloads_output_file <- "../output/gbif_downloads_list.tsv"
```

This text file will contain all informations about triggered downloads from GBIF.

# Define taxa

We are interested in all Chordata at this moment and there is no a unique worth-of-thrust checklist for all UKOTs.:

```{r define_taxa}
taxa_input_file <- NA
```

## Upload list of countries

A list of countries has been made by Oliver Scott in his [142--UKOT--horizon-scanning repository](https://github.com/sacrevert/142--UKOT--horizon-scanning). We import it as a data.frame

```{r upload_country_list}
country_list_file <- "../data/CaribbeanOT_links_GATS_and_PA_withISOs_v1.csv" # for offline use
country_list_file <- "https://raw.githubusercontent.com/sacrevert/142--UKOT--horizon-scanning/master/data/CaribbeanOT_links_GATS_and_PA_withISOs_v1.csv" # for online use
country_df <- read_csv(file = country_list_file)
```

Extract UKOTs with their corresponding country code:

```{r extract_countries}
countries_reporter2iso <- country_df %>% 
  distinct(reporter2iso, reporter) %>% 
  filter(!is.na(reporter2iso)) %>%
  rename(country = reporter,
         countryCode = reporter2iso)
paste("UKOTs:", paste(countries_reporter2iso$country, collapse = ","))
paste("UKOT country  codes: ", paste(countries_reporter2iso$countryCode,
      collapse = ","))
```

# Download GBIF occurrences

We spin up a download of occurrence data related to UKOT countries:

```{r trigger_gbif_download_UKOT, eval = FALSE}
gbif_download_key_UKOT <- occ_download(
  paste0("country = ", 
         paste(countries_reporter2iso$countryCode, collapse = ",")),
  curlopts = list(verbose = TRUE))
```

**Note**: you need a valid GBIF account to trigger a download. You can save your credentials in environment file of R (`.Renviron` or `.Renviron.site`) as follows:

```{r show_GIBF_credentials, eval = FALSE}
gbif_user = "my_username"
gbif_pwd = "my_password"
gbif_email = "my_email_address"
```

```{r offline_import_key, include = FALSE}
gbif_download_key_UKOT <- "0008649-180412121330197"
```

Check status of download:

```{r retrieve_status_download}
metadata_UKOT <- occ_download_meta(key = gbif_download_key_UKOT)
metadata_UKOT$key
metadata_UKOT$status
```

In case status is `RUNNING` or `PREPARING`, no worry: it takes time to retrieve so many observations. Take a coffie or a cup of the or check your e-mails. Run the code above within a few minutes to check the status of your request. Once `metadata_UKOT$status` is `COMPLETED`, you are ready for downloading.

Within [trias project](https://github.com/trias-project) we are developing a packge, [`trias`](https://github.com/trias-project/trias), where we wrote a function to update a list of all triggered GBIF downloads. In this way we have a handful overview of all GBIF keys and details related to every download. In our case `input_checklist` is `NA`.

```{r _update_downloads}
update_download_list(
  file = downloads_output_file, 
  download_to_add = gbif_download_key_UKOT, 
  input_checklist = taxa_input_file)
```

## Import downloads to R

It can be useful to import occurrence data directly in R. 
You can do it by GBIF download key. In our case, for occurrence related to UKOT countries:

```{r import_ocurrence_from_gbif_key_UKOT}
gbif_key <- metadata_UKOT$key
file_path <- "../output/"
occ <- occ_download_get(key = gbif_key, overwrite = TRUE, path = file_path)
file_to_extract <- "occurrence.txt"
unzip(zipfile = occ, files = file_to_extract)
file_name <- paste0(file_path, gbif_key, "_", file_to_extract)
file.rename(file_to_extract,
            file_name)
occurrence_UKOT <- read_delim(file_name, 
                         "\t", escape_double = FALSE, trim_ws = TRUE)
```


# Select species from checklists

## IAS Caribbean database

Import a reduced version of the IAS Caribbean database as provided by Oliver Scott.

```{r import_iasDB}
iasDb_file <- "../data/tabula-IASCaribDbReduced_FINAL.csv" # for offline use
iasDb_file <-"https://raw.githubusercontent.com/sacrevert/142--UKOT--horizon-scanning/master/data/tabula-IASCaribDbReduced_FINAL.csv" # for online use
iasDb <- read_csv(file = iasDb_file)
```

We clean it a little up
```{r cleaned_up_version_iasDB}
iasDb %<>% 
  janitor::clean_names() %>% # standardize column names
  rename(species = species_name) %>% #rename column species_name as species
  slice(2:n()) %>% # remove first row containing subtitles
  mutate(
    exotic_in = str_replace(exotic_in,  # change Turks-Caicos to ISO name
                            pattern = "Turks-Caicos", 
                            replacement = "Turks and Caicos Islands"),
    invasive_in = str_replace(invasive_in,
                              pattern = "Turks-Caicos",
                              replacement = "Turks and Caicos Islands")) %>%
  distinct() %>% # remove duplicates
  mutate(checklist = "Caribbean_IAS_db") # add checklist source tag
head(iasDb)
```


Select vertebrates by column `organism_type`

```{r filter_vertebrates}
iasDb_vert_invert <- iasDb %>% filter(str_detect(organism_type, pattern = "ertebrate"))
head(iasDb_vert_invert)
```

## GRIIS

We have also a partial checklist of invasive species for Anguilla  and Bermuda from [GRIIS](http://www.griis.org/) as a `.csv` file:

```{r}
griis_AI_BM <- read_csv2(
  "../data/export_griis_Anguilla_Bermuda.csv") # for offline use
griis_AI_BM %<>% janitor::clean_names()
head(griis_AI_BM)
```

We select the Chordata via GBIF taxonomy backbone:

```{r select_Chordata_via_GBIF_backbone}
alien <- griis_AI_BM %>% distinct(origin) %>% pull(origin) # origin = "Alien"
griis_AI_BM <-  griis_AI_BM %>% 
      rowwise() %>%
      do_(interp(~ as.data.frame(name_usage(name = .$species,
                                                     return = "data")))) %>%
  distinct(species, phylum) %>%
  mutate(checklist = "GRIIS",
         origin = alien) %>% 
  filter(!is.na(species))
griis_AI_BM_vert_invert <- griis_AI_BM %>% 
  filter(str_to_lower(phylum) == str_to_lower("Chordata")) %>% 
  select(species, checklist, origin, phylum)
griis_AI_BM_vert_invert
```

Doing it, we have removed the following species from GRIIS checklist:

```{r not_chordata_from_griis}
anti_join(griis_AI_BM %>% 
            filter(!is.na(species)) %>%
            filter(!is.na(phylum)) %>%
            distinct(), 
          griis_AI_BM_vert_invert,
          by = c("species", "checklist", "phylum"))
```

## Creation of a unified checklist

We can merge now the two checklists for alien vertebrates:

```{r merge_lists_AS}
unified_checklist_vert_invert <- iasDb_vert_invert %>% 
  full_join(griis_AI_BM_vert_invert, by = c("species", "checklist"))
head(unified_checklist_vert_invert)
```

A fast check shows the overlapping degree of the two checklists:

```{r}
print(paste("Number of vertebrates in IAS Caribbean database:",
            nrow(iasDb_vert_invert)))
print(paste("Number of vertebrates in GRIIS checklist:",
            nrow(griis_AI_BM_vert_invert)))
print(paste("Number of vertebrates in both databases:",
            nrow(iasDb_vert_invert %>% 
                   filter(species %in% 
                            (griis_AI_BM_vert_invert %>% pull(species))))))
```

# Number of observations of alien species in each UKOT

We use column `species` from `unified_checklist_vert_invert` for searching observations in occurrence data.frame `occurrence_UKOT`. We are interested in counting how many observations are for each species in each UKOT:

```{r}
occ_vert_invert <- occurrence_UKOT %>% 
  filter(species %in% 
           unified_checklist_vert_invert$species) %>%
  group_by(species, countryCode) %>% 
  count() %>% ungroup() %>% 
  spread(key = countryCode, value = n) %>%
  right_join(unified_checklist_vert_invert,
            by = "species")
occ_vert_invert
```

Save file in a text file

```{r save_tsv_file}
write_tsv(occ_vert_invert, 
          path = "../output/ias_UKOT_vertebrates_invertebrates.tsv")
```

Next time you work on these data, you can just directly load this file by:

```{r load_tsv_file, eval = FALSE}
occ_vert_invert <- 
  read_tsv(file = "../output/ias_UKOT_vertebrates_invertebrates.tsv")
```

# Some considerations about checklist completeness

In IAS Caribbean database there are two columns containing information about status of invasion: `invasive_in` and `exotic_in`. For example, `Passer domesticus` is considered exotic in 

```{r example_exotic_iasDB}
iasDb_vert_invert %>% 
  filter(species == "Passer domesticus") %>%
  select(exotic_in)
```

and invasive in 

```{r}
iasDb_vert_invert %>% 
  filter(species == "Passer domesticus") %>%
  select(invasive_in)
```

In GRIIS checklist for Anguilla and Bermuda all species share the same `origin` column value: `Alien`, so they have been all considered alien species by experts. Although occurrence data cannot negate such judge, they can still be used to find new potential alien species which are still not present in any alien species checklist.

First of all, we select species and relative observation counts for each UKOT where not NA and gather them in a tidy table (country code as column):

```{r species_country_distr}
occ_vert_invert_tidy <- occ_vert_invert %>% 
  select(species, one_of(countries_reporter2iso$countryCode)) %>%
  gather(key = "countryCode", value = "n", countries_reporter2iso$countryCode) %>%
  filter(!is.na(n)) %>% 
  distinct() %>% 
  left_join(countries_reporter2iso, by = "countryCode")
occ_vert_invert_tidy
```

Do GBIF occurrence data contain new information about alien (or exotic) species in UKOTs?

The first simplest check is the following: find species in `occ_vert_invert_tidy` which are not present in `unified_checklist`:

```{r species_not_in_unified_checklist}
occ_vert_invert_tidy %>% distinct(species) %>% filter(! species %in% unified_checklist_vert_invert$species)
```

Let's go more in details considering the occurrence data in each UKOT and the geographical information contained in checklists.

```{r}
gbif_invasive <- map_df(
  countries_reporter2iso$country,
  function(x) 
    inner_join(occ_vert_invert_tidy %>% 
                 filter(country == x),
               unified_checklist_vert_invert %>% 
                 filter(!str_detect(exotic_in, x) &
                          !str_detect(invasive_in, x) &
                          is.na(origin)),
               by = "species"))
```

Scrolling these lists per each country we can see that GBIF occurrences show that the unified checklist made by joining Caribbean IAS database and  Anguilla and Bermuda's GRIIS checklist are not udpated. We save them as tsv file:

```{r GBIF_occ_not_in_checlists}
write_tsv(gbif_invasive, 
                 path = "../output/gbif_invasive_not_in_checklists.tsv")
```

The unified checklist, completed by these additional species found by GBIF, has been sent to local experts leaded by Tim Adriaens for a check. Their revision was helpful for removing synonyms, correcting habitat information, etc.
We used a Google spreadsheet for doing it. 
We can upload the most updated version using code here below:

```{r get_most_udpated_longlist}
my_sheets <- gs_ls()
google_ssheet <- my_sheets %>% 
  filter(str_detect(my_sheets$sheet_title, 
                    pattern = "ias_UKOT_vertebrates_invertebrates"))
google_ssheet %<>%
  filter(updated == max(google_ssheet$updated)) # take the most updated version
unified_checklist_vert_invert <- gs_key(google_ssheet$sheet_key) %>% gs_read(ws = "LONGLIST")
unified_checklist_vert_invert %<>% clean_names()
```


# Is the "Horizon" already here? A case study

Based on file `Horizon Scanning_20180516144937707_CABI_verts_FloridaMexicoCuba.csv` from CABI we can have an idea of the possible future invasive species at **Cayman Islands**.

```{r import_file_horizon_scan}
horizon_KY <- read_csv("../data/Horizon Scanning_20180516144937707_CABI_verts_FloridaMexicoCuba.csv", skip = 5) %>%
  janitor::clean_names() %>%
  rename(species = preferred_scientific_name)
horizon_KY
```

Which of them have been already observed in the Cayman Islands based on GBIF occurrences?

```{r already_present_KY_occ_based}
new_introductions_KY_occ <- horizon_KY %>% 
  inner_join(occ_vert_invert_tidy %>% filter(countryCode == "KY"),
             by = "species")
new_introductions_KY
```

Which of them have been labelled as invasive or exotic in the Caribbean Invasive Alien Species Database?

```{r already_present_KY_checklist_based_invasive}
new_introductions_KY_checklist <- horizon_KY %>% 
  inner_join(unified_checklist_vert_invert %>% 
  filter(str_detect(invasive_in, "Cayman Islands") |
           str_detect(exotic_in, "Cayman Islands")),
  by = "species")
new_introductions_KY_checklist
```

Therefore, we can conclude that the following species are not just on the horizon, but they are unforunately present at the Cayman Islands:

```{r already_present_KY_final}
new_introductions_KY <- new_introductions_KY_checklist %>% 
  select(species, checklist) %>% 
  bind_rows(new_introductions_KY_occ %>% 
              select(species, n) %>% 
              mutate(checklist = "GBIF_occurrence")) %>% 
  distinct(species, checklist, n)
new_introductions_KY
```

Now we combine the horizon scan of Caribbean UKOTs against EU horizon scan countries (Roy et al.)

```{r horizon_scan_UKOTs_Europe}
horizon_UKOTs_EU <- read_csv("../data/EU_HorizonScanning.csv",
                             skip = 1) %>%
  janitor::clean_names()
horizon_UKOTs_EU
```

Which of them have been already observed in the UKOTs based on GBIF occurrences?

```{r already_present_KY_occ_based}
new_introductions_UKOTs_occ <- horizon_UKOTs_EU %>% 
  inner_join(occ_vert_invert_tidy,
             by = "species")
new_introductions_UKOTs_occ %>% select(species, country, n)
```

